Supplementary Information: Quantifying the pathways to life using assembly spaces

### 2021-05-21 papers/assembly/I/main.tex
Flesh out text for the assembly index section

\newcommand{\upset}[1]{#1\!\!\uparrow}
\newcommand{\downset}[1]{#1\!\!\downarrow}
\newcommand{\N}{\mathbb{N}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\Ini}[1]{\textrm{Ini}(#1)}
\newcommand{\Term}[1]{\textrm{Term}(#1)}
\newcommand{\A}{\mathcal{A}}
\newcommand{\powerset}{\mathcal{P}}
\newcommand{\pathway}[1]{\langle#1\rangle}
\newcommand{\R}{\mathbb{R}}

@@

\begin{definition}\label{def:size}
    Let $A$ be an assembly space.
    The \textbf{size} of $A$ is the cardinality of $T(A)$, denoted $|A|$.
\end{definition}

\begin{definition}\label{def:downset}
    Let $(A, \B_A)$ be an assembly space and let $x$ be a site in $A$.
    The \textbf{down set} of $x$, denoted $\downset{x}$ in $A$ is the site completion of all pathways from $\B_A$ terminating at $x$.
\end{definition}

\begin{definition}\label{def:finiteness}
    Let $A$ be an assembly space and let $x$ be a site in $A$.
    We say that $x$ is \textbf{finite} in $A$ if $\downset{x}$ in $A$ has finite size.
\end{definition}

\begin{definition}\label{def:index}
    Let $A$ be an assembly space and let $x$ be a finite site in $A$.
    The \textbf{assembly index} of $x$ is the minimal size of all rooted assembly subspaces of $A$ containing $x$, and is denoted $c_A(x)$.
    This can be written $c(x)$ when the relevant assembly space is clear from context.
\end{definition}

\begin{theorem}\label{def:index-existence-uniqueness}
    Let $A$ be an assembly space and let $x$ be a finite site in $A$.
    Then the assembly index $c_A(x)$ is exists and is unique.
\end{theorem}
\begin{proof}
    \TODO{Complete: likely requires Zorn's lemma, but maybe not...}
\end{proof}

\subsection{Bounds on the Assembly Index}\label{subsec:assembly-index-bounds}

\begin{theorem}\label{cor:subspace-index-upper-bound}
    Let $(A, \B_A)$ be an assembly space and let $(B, \B_A)$ be a rooted assembly subspace of $A$.
    For every finite $b \in S(B)$, the assembly index of $b$ in $B$ is at least the assembly index of $b$ in $A$.
    That is, $c_B(b) \ge c_A(b)$ for all $b \in S(B)$
\end{theorem}
\begin{proof}
    \TODO{Complete}
\end{proof}

\begin{theorem}\label{thm:image-lower-bound}
    If $f: A \rightarrow B$ is an assembly map, then $c_{f(A)}(f(x)) \le c_A(x)$ for all finite $x \in S(A)$.
    Furthermore, if $f$ is an embedding, then $c_{f(A)}(f(x)) = c_A(x)$.
\end{theorem}
\begin{proof}
    \TODO{Complete}
    % Suppose $x \in S(A)$ is finite, and $Z \subseteq A$ is a rooted with $s \in S(Z)$ and $|Z| = c_A(x)$.
    % Restricting $f$ to $Z$ provides an assembly map $f^* : Z \rightarrow B$ such that $f(Z)$ contains $f(x)$, and is rooted in $B$.
    % As such, $c_{f(A)}(f(x))$ can be no larger than $|f(Z)|$, so $c_{f(A)}(f(x)) \le c_A(x)$.
    %
    % Furthermore, suppose that $f$ is an embedding, and consider any $Z' \subseteq f(A)$ rooted and containing $f(x)$.
    % The preimage of $Z'$ under $f$ is then a rooted assembly subspace of $A$ containing $x$, so $c_A(x)$ can be no larger than $|f^{-1}(Z)|$.
    % Together with the previous result, we then have $c_{f(A)}(f(x)) = c_A(x)$.
\end{proof}

\begin{corollary}\label{cor:lowerset-index}
    Let $A$ be an assembly space and $x \in S(A)$ be finite.
    Then $c_A(x) = c_{x\downarrow}(x)$.
\end{corollary}
\begin{proof}
    \TODO{Complete}
    % There is a natural embedding $i : \downset{x} \rightarrow A$, so $c_A(x) = c_{x\downarrow}(x)$ for all finite $x \in S(A)$, as per \cref{thm:image-lower-bound}.
\end{proof}

\subsection{Computability and Algorithms}

\begin{theorem}\label{thm:computable}
    If $A$ is an assembly space and $x \in S(A)$ is finite, then $c(x)$ is computable.
\end{theorem}
\begin{proof}
    \TODO{Complete}
    % Every rooted assembly subspace with minimal size and containing $x$ is a minimal, rooted assembly subspace of $\downset{x}$.
    % Since $x$ is finite, $\downset{x}$ is finite; that is $\downset{x}$ has finitely many transitions.
    % This, together with the fact that $\pi(\sigma(t))$ and $\pi(\tau(t))$ are finite for all $t \in T(A)$, ensures that there areonly finitely many assembly subspaces of $\downset{x}$.
    % As such, the set of all rooted assembly subspaces is computable; simply enumerate them.
    % The size of each subspace is, of course, computable as one need simply count the transitions in the subspace.
    % Finally, finding the minimum value in a finite set of natural numbers is computable.
    % Thus, $c(x)$ is computable.
\end{proof}

\TODO{Typeset split-branch algorithm}

\section{Assembly Examples}

\TODO{Come up with a few fun examples.}

>>>>

\newcommand{\upset}[1]{#1\!\!\uparrow}
\newcommand{\downset}[1]{#1\!\!\downarrow}
\newcommand{\N}{\mathbb{N}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\SB}{\mathcal{S}}
\newcommand{\Ini}[1]{\textrm{Ini}(#1)}
\newcommand{\Term}[1]{\textrm{Term}(#1)}
\newcommand{\A}{\mathcal{A}}
\newcommand{\powerset}{\mathcal{P}}
\newcommand{\pathway}[1]{\langle#1\rangle}
\newcommand{\R}{\mathbb{R}}

@@

With the core formalism established, we can now present the concept of the \textit{assembly index}.
As we have seen, within the context of an assembly space, the rooted assembly spaces are partially ordered.
This allows us to inquire about the minimal, rooted assembly subspaces which contain some desirable site.
Each of these subspaces will consist of some, possibly infinite, number of transitions.
We refer to the number of transitions in an assembly space as the \textit{size}.
\begin{definition}\label{def:size}
    Let $A$ be an assembly space.
    The \textbf{size} of $A$ is the cardinality of $T(A)$, denoted $|A|$.
\end{definition}
In the running ligation model, $\A$, the size of the entire space is $|\A| = 4$ as there are four transitions.
Similarly, the minimal, rooted subspace of $\A$ containing $AA$ has size $1$; it contains only one transition, $\alpha$.

We can think of the size of a minimal, rooted subspace containing a given site as counting the number of steps required to assemble that site in a particular way.
However, it is entirely possible that there is no finite way assembly an site.
For example, consider a polymer model which consists of combining monomers $A$ and $B$ into polymers of arbitrary length via pairwise joining operations.
The space of all polymers naturally extends to infinite-length polymers such as $ABABA\ldots$ which require an infinite number of steps to assemble.
While the abstract assembly space may well contain such sites, we wish to restrict ourselves to sites which can be assembled in finitely many steps.
We say that such a site if $\textit{finite}$.
\begin{definition}\label{def:finiteness}
    Let $A$ be an assembly space and let $x$ be a site in $A$.
    We say that $x$ is \textbf{finite} in $A$ there exists a rooted assembly subspace of $A$ which contains $x$ as a site.
\end{definition}
Every site in our running example space, $\A$ is finite.
In the polymer model, only sites representing polymers with finitely many monomers are finite.
If we are presented with such a finite site, we can count the size of every rooted assembly subspaces containing that site.
The least size of these subspaces is the \textit{assembly index} of the site.
\begin{definition}\label{def:index}
    Let $A$ be an assembly space and let $x$ be a finite site in $A$.
    The \textbf{assembly index} of $x$ is the minimum size of all rooted assembly subspaces of $A$ containing $x$, and is denoted $c_A(x)$.
    This can be written $c(x)$ when the relevant assembly space is clear from context.
\end{definition}
In this definition, we do not restrict ourselves specifically to minimal subspaces containing the site $x$.
However, in effect we are.
If we were to consider the size of a non-minimal, rooted subspace some site $x$, we'd always be able to find some minimal, rooted subspace which contains $x$ and has at most the same size.
Since we are looking specifically for the \textit{minimum} size of the subspaces, we need to confirm that such a value even exists.
It does and is in fact unique.
\begin{theorem}\label{def:index-existence-uniqueness}
    Let $A$ be an assembly space and let $x$ be a finite site in $A$.
    Then the assembly index $c_A(x)$ is exists and is unique.
\end{theorem}
\begin{proof}
    \TODO{Complete: likely requires Zorn's lemma, but maybe not...}
\end{proof}

To conclude this section, we present the assembly index of every site of in our ligation model $\A$ together with a minimal, rooted assembly subspace containing that site.

\TODO{Complete Figure}

In the next section, we will consider methods for bounding the assembly index of a site.

\subsection{Computability of and Bounds on the Assembly Index}\label{subsec:assembly-index-bounds}

The first task, when presented with an assembly space, is to determine the assembly index of the sites in the space.
In practical terms, this requires an algorithm for computing the assembly index relative to some space.
In this work, we do not present a general algorithm to do so, though a simple depth-first search of the space may suffice in some cases.
Rather, we first show that the assembly index is in fact computable.
\begin{theorem}\label{thm:computable}
    If $A$ is an assembly space and $x \in S(A)$ is finite, then $c(x)$ is computable.
\end{theorem}
\begin{proof}
    \TODO{Complete}
\end{proof}
Of course, just because the value is computable does not mean that it can be computed in reasonable time.
In fact, the general problem is likely to be \texttt{NP}-hard.
In situtations where direct computation is infeasible, we can resort to bounding the value of the assembly index by computing the index in an analogous space.
The first of the bounds we present is an upper bound determined by computing the assembly index in a subspace containing the site.
\begin{theorem}\label{cor:subspace-index-upper-bound}
    Let $(A, \B_A)$ be an assembly space and let $(B, \B_A)$ be a rooted assembly subspace of $A$.
    For every finite $b \in S(B)$, the assembly index of $b$ in $B$ is at least the assembly index of $b$ in $A$.
    That is, $c_B(b) \ge c_A(b)$ for all $b \in S(B)$
\end{theorem}
\begin{proof}
    \TODO{Complete}
\end{proof}

\TODO{Example and Figure}

The second bound we establish is a lower bound determined by mapping the space in question into a space within which computing the assembly index may be simpler.
\begin{theorem}\label{thm:image-lower-bound}
    If $f: A \rightarrow B$ is an assembly map, then $c_{f(A)}(f(x)) \le c_A(x)$ for all finite $x \in S(A)$.
    Furthermore, if $f$ is an embedding, then $c_{f(A)}(f(x)) = c_A(x)$.
\end{theorem}
\begin{proof}
    \TODO{Complete}
\end{proof}

\TODO{Example and Figure}

Our final bound is a special case of the lower bound derived from an assembly map; however, it represents an upper bound.
Given an assembly space with no directed loops, which we refer to as \textbf{acyclic}, we can form an analogous assembly space which has a tree-like structure by duplicating sites that participate in more than one transition.
We refer to such a space as the \textit{split-branch} representation of the space.
\begin{definition}\label{def:split-branch}
    Let $A$ be an acyclic assembly space.
    The \textbf{split-branch} representation of $A$ is $\ldots$.

    \TODO{Doug is leaving this for Stu since it's his jam.}
\end{definition}

There is a natural assembly map from the split-branch representation of a space to the space itself.
As such, \Cref{thm:image-lower-bound} applies, and we can produce an upper bound on the assembly index of sites in the original space.
\begin{corollary}\label{cor:split-branch}
    Let $A$ be a \red{adjectives} assembly space and let $\SB(A)$ be the split-branch representation of $A$, and let $f : \SB(A)$ such that $\ldots$.
    For any site $x$ in $A$, the assembly index $c_A(x) \le \min_{x' \in S(f)^{-1}(x)}c_{\SB(A)}(x')$.
\end{corollary}
\begin{proof}
    \TODO{Stu to complete? Or Doug can with a workable \Cref{def:split-branch}.}
\end{proof}

\TODO{Typeset algorithm}
\red{I need a PDF version of the previous version of the supplement. The formatted algorithm not rendering correctly for me in Word.}
